---
title: "Introduction to time series analysis"
author: "Boyko Amarov"
date: "10/12/2021"
---

# Time series data in R

## Examples of time series?

- temperature (weather), e.g. daily average temperature
- monthly rainfall
- yearly sales of bread in some store
- time to travel to work (monthly, yearly, daily)

```{r}
# install.packages(c("xts", "ggplot2"))
# Extended Time Series
library(xts)
library(ggplot2)

bitcoins <- read.csv(
  'https://s3.eu-central-1.amazonaws.com/sf-timeseries/data/bitcoin2017-06-01-2017-11-13.csv', 
  stringsAsFactors = FALSE
)

## Creates an extended time series object
bc_ts <- xts(bitcoins$close, order.by = as.Date(bitcoins$Index))
```


## Difference from cross-section (mock example)


```{r}
cross_section <- data.frame(
  person = c(1, 2, 3, 4, 5, 6),
  party = c("A", "B", "C", "C", "B", "C")
)

cross_section1 <- cross_section[c(2, 1, 3, 5, 4, 6), ]

table(cross_section$party) / nrow(cross_section)
```


- Plot the time series

```{r}
plot(bc_ts)
```

We'll focus on three properties of time series:

1. Level (Niveau): measured by the mean of the time series
2. Fluctuations (variation): measured by the variance of the time series (standard deviation)
3. Dependencies between the measurements in the time series: measured by correlations


- Compute the average exchange rate over the whole observation period

```{r}
## Using the mean function
mean(bc_ts)
```
- Compute the average exchange rate for June 2017

```{r}
## Subset the time series, then apply the mean function
mean(bc_ts["2017-6"])
```

- Compute the average exchange rate for October 2017

```{r}
## Same as for October 2017
mean(bc_ts["2017-10"])
```

- Compute the average exchange rate before (and including) and after October 2017

```{r}
## Use a slash to select everything from the start of the time series
## to the end of October
mean(bc_ts["/2017-10"])
```

```{r}
## Average exchange rate after the end of october
mean(bc_ts["2017-11/"])
```


- Compute the standard deviation for the four periods

## Overall standard deviation

```{r}
sd(bc_ts)
```

```{r}
## Subset the time series, then apply the sd function (which computes the standard deviation)
sd(bc_ts["2017-6"])
```

```{r}
## Same as for October 2017
sd(bc_ts["2017-10"])
```

- Create a scatterplot for the exchange rate time series and its first lag

$$
y_{1}, y_{2}, y_{3}, \ldots, y_{T}\\
y_{t}, t = 1,\ldots, T
$$
In the bitcoins example $T = 117$. $y_1 = 2475.248, y_2 = 2688.172, \ldots, y_{117} = 6535.948$.

Looking from Nov 13th 2017: what was the value on the previous trading day? Answer
by simply looking in the table: $y_{117 - 1} = y_{116} = 6622.517$. 

$$
y_{t}, t = 1,\ldots, T\\
y_{t - 1}: \text{ first lag of } y_{t}\\
y_{t - 2}: \text{ second lag of } y_{t}
$$

```{r}
bc_ts_l1 <- lag(bc_ts)

tmp <- cbind(bc_ts, bc_ts_l1)

ggplot(data = tmp, aes(x = bc_ts_l1, y = bc_ts)) +
  geom_point() +
  labs(
    x = expression(y[t-1]),
    y = expression(y[t])
  )
```
